---
title: "02_ES_derivedEC_comparison"
author: "Sera Choi"
date: "12/07/2021"
output: html_document
---
```{r, message=FALSE, warning=FALSE}
# Load library
library(Seurat)
library(ggplot2)
library(dplyr)
library(here)
library("scales")
library(ggsci)
library(gtools)
library(future)

mypal <- pal_npg("nrc", alpha = 0.7)(10)
mypal2 <-pal_tron("legacy", alpha = 0.7)(7)
mypal3 <- pal_lancet("lanonc", alpha = 0.7)(9)
mypal4 <- pal_simpsons(palette = c("springfield"), alpha = 0.7)(16)
mypal5 <- pal_rickandmorty(palette = c("schwifty"), alpha = 0.7)(6)
mypal6 <- pal_futurama(palette = c("planetexpress"), alpha = 0.7)(5)
mypal7 <- pal_startrek(palette = c("uniform"), alpha = 0.7)(5)
mycoloursP<- c(mypal, mypal2, mypal3, mypal4, mypal5, mypal6, mypal7)
show_col(mycoloursP, labels =F)

# Set maximum size of the object 
options(future.globals.maxSize = 50000*1024^2)
```

# ES-derived EC vs. Young Adult Brain EC

This section finds similarities and differences between ES-derived ECs and Young Adult Brain ECs. PC1 represents differences within McCracken's dataset. 

```{r}
# load data
ec <- readRDS("/Users/serachoi/Documents/Edinburgh/MScProject/Project/data/processed/ES_EC_integrated.RDS")

# Visualisation
DimPlot(ec, reduction = "pca", group.by = "orig.ident")
DimPlot(ec, reduction = "umap", group.by = "orig.ident")

# PCA analysis 
DimHeatmap(ec, dims = 1, cells = 500, balanced = TRUE, assays = "integrated")
print(ec[["pca"]], dims = 3, nfeatures = 20)
```


# Differential gene expression analysis 

## Enhanced in ES-derived ECs 

Of 3,241 DEGs, all were significant (adjusted p_value < 0.05). After applying more stringent thresholds, 2994 remained. 

```{r, fig.width= 10}
DefaultAssay(ec) <- "RNA"
Idents(ec) <- "orig.ident"

es.markers <- FindMarkers(ec,
                           ident.1 = "McCracken_H9_Day8",
                           ident.2 = "Seeker_BECs",
                           only.pos = TRUE, 
                           min.pct = 0.25, 
                           logfc.threshold = 0.25,
                           test.use = "MAST")
es.markers
write.csv(es.markers, "/Users/serachoi/Documents/Edinburgh/MScProject/Project/docs/es_up_unfiltered.csv")

es.markers_sig <- subset(es.markers, es.markers$p_val_adj < 0.05 )

es.markers_stringent <- subset(es.markers, es.markers$p_val_adj < 0.05 & es.markers$pct.1 > 0.25 & es.markers$pct.2 < 0.25)
write.csv(es.markers_stringent, "/Users/serachoi/Documents/Edinburgh/MScProject/Project/docs/es_up.csv")

# Dotplot
pdf("/Users/serachoi/Documents/Edinburgh/MScProject/Project/outs/es-derived/dot_es.pdf", height = 3.7, width = 10)
dittoDotPlot(object = ec, 
             vars = rownames(es.markers_stringent)[1:20], 
             group.by = "orig.ident", 
             scale = F,
             min.color = "navyblue",
             max.color = "orangered",
             size = 10,
             assay = "RNA") + theme_light() +
  theme(axis.text = element_text(size=13), axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))
dev.off()
```

### Ribosomal gene 

```{r}
# ribo genes 
rpl <- PercentageFeatureSet(ec, pattern = "^RPL", assay = "RNA")
rps <- PercentageFeatureSet(ec, pattern = "^RPS", assay = "RNA")
ec[["percent.ribo"]] <- rpl+rps

# Plot
pdf("/Users/serachoi/Documents/Edinburgh/MScProject/Project/outs/es-derived/ribo.pdf", height = 5, width = 6)
VlnPlot(ec, features = "percent.ribo", group.by = "orig.ident", cols = mycoloursP[1:20])
dev.off()
```

## GSEA with MSigDB Hallmarks 

```{r, fig,width = 15, fig.height=6}
## Format data
# Return true/false where gene symbols in grch38 matches that in list of DEGs
idx <- grch38$symbol %in% rownames(es.markers)
# Subset for rows in grch38 that matches DEGs 
ids <- grch38[idx,]

# The gene names can map to more than one Ensembl ID (some genes change ID over time), 
# so we need to remove duplicate IDs prior to assessing enriched GO terms
non_duplicates <- which(duplicated(ids$symbol) == FALSE)
ids <- ids[non_duplicates, ] 

# Add gene column in DEG list 
es.markers["gene"] <- rownames(es.markers)

# Merge DEGs with subset of table from grch38
res_ids <- inner_join(es.markers, ids, by=c("gene" ="symbol")) 

# Create dataset containing significant genes 
sig_ids <- subset(res_ids, p_val_adj < 0.05)

# remove NAs
res_entrez <- subset(sig_ids, entrez != "NA")

# Remove duplicates 
res_entrez <- res_entrez[which(duplicated(res_entrez$entrez) == F), ]

# Extract name and fold-changes
foldchanges <- res_entrez$avg_log2FC

# Name each fold change with the corresponding Entrez ID
names(foldchanges) <- res_entrez$entrez

# Order in decreasing order
foldchanges <- sort(foldchanges, decreasing = TRUE)

## Load GMT file downloaded from MSigDB website 
h <- read.gmt("/Users/serachoi/Documents/Edinburgh/MScProject/Project/data/MSigDB_H")

# Hallmarks
msig_h_es <- GSEA(foldchanges, 
               TERM2GENE=h,
               scoreType = "pos",
               minGSSize = 10,
               pvalueCutoff = 1,
               pAdjustMethod = "BH")

msig_H_result_es <- data.frame(msig_h_es)
msig_H_tidy_es <- subset(msig_H_result_es, subset = msig_H_result_es$p.adjust<0.6)

#Barplot
ggplot(msig_H_tidy_es, aes(reorder(Description, NES), NES)) + geom_col(aes(fill=p.adjust< 0.2)) + theme_minimal() + coord_flip() + theme(axis.text = element_text(size = 11))
```

## GSEA with MSigDB Gene ontology

```{r, fig.width=10, fig.height=5}
## Load GMT file downloaded from MSigDB website 
c5 <- read.gmt("/Users/serachoi/Documents/Edinburgh/MScProject/Project/data/MSigDB_C5")

# Hallmarks
msig_c5_es <- GSEA(foldchanges, 
               TERM2GENE=c5,
               scoreType = "pos",
               minGSSize = 10,
               pvalueCutoff = 1,
               pAdjustMethod = "BH")

msig_C5_result_es <- data.frame(msig_c5_es)

msig_C5_tidy_es <- subset(msig_C5_result_es, subset = msig_C5_result_es$p.adjust<0.6)

# Remove GOBP and _ for prettier label for graph 
pathway <- rownames(msig_C5_tidy_es)
pathway <- str_remove_all(pathway, "GOBP_")
pathway <- str_replace_all(pathway, "_", " ")
pathway <- tolower(pathway)
msig_C5_tidy_es$Description <- pathway

#Barplot
pdf("/Users/serachoi/Documents/Edinburgh/MScProject/Project/outs/es-derived/GO_es.pdf", height = 8, width = 13)
ggplot(msig_C5_tidy_es[1:20,], aes(reorder(Description, NES), NES)) + geom_col(aes(fill=p.adjust< 0.2)) + theme_light() + coord_flip() + theme(axis.text = element_text(size = 18))+ scale_fill_manual(values = c(mycoloursP[5],mycoloursP[4])) + labs(x="", y="NES")
dev.off()
```


## Enhanced in mature ECs 

Of 1362 DEGs, all were shown to be significant. After applying stringent thresholds, 371 genes were up-regulatd in mature ECs compared to ES-derived ECs. 

```{r, fig.width=10, fig.height=6}
DefaultAssay(ec) <- "RNA"
Idents(ec) <- "orig.ident"

ec.markers <- FindMarkers(ec,
                           ident.1 = "Seeker_BECs",
                           ident.2 = "McCracken_H9_Day8",
                           only.pos = TRUE, 
                           min.pct = 0.25, 
                           logfc.threshold = 0.25,
                           test.use = "MAST")
ec.markers
write.csv(ec.markers, "/Users/serachoi/Documents/Edinburgh/MScProject/Project/docs/ec_up_unfiltered.csv")

ec.markers_sig <- subset(ec.markers, ec.markers$p_val_adj < 0.05 )

ec.markers_stringent <- subset(ec.markers, ec.markers$p_val_adj < 0.05 & ec.markers$pct.1 > 0.25 & ec.markers$pct.2 < 0.25)
write.csv(ec.markers_stringent, "/Users/serachoi/Documents/Edinburgh/MScProject/Project/docs/ec_up.csv")


# BarPlot
ec.markers_stringent$gene <- rownames(ec.markers_stringent)
ec.markers_stringent$log10_FDR <- -log10(ec.markers_stringent$p_val_adj)
ggplot(data = ec.markers_stringent[1:20,], aes(x = reorder(gene, -avg_log2FC), y = avg_log2FC, fill = log10_FDR)) + geom_bar(stat="identity")  + theme_light()+ theme(axis.text = element_text(size=13), axis.text.x = element_text(angle = 45, vjust = 0.5, hjust=0.4), legend.text = element_text(size = 12), legend.title = element_text(size = 13), text = element_text(size = 15)) + scale_fill_continuous(type="viridis")

# Dotplot
pdf("/Users/serachoi/Documents/Edinburgh/MScProject/Project/outs/es-derived/dot_ec.pdf", width = 10, height = 3.7)
dittoDotPlot(object = ec, 
             vars = rownames(ec.markers_stringent)[1:20], 
             group.by = "orig.ident", 
             scale = F,
             min.color = "navyblue",
             max.color = "orangered",
             size = 10,
             assay = "RNA") + theme_light() +
  theme(axis.text = element_text(size=13), legend.text = element_text(size = 12), axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) 
dev.off()
```

## GSEA with MSigDB Hallmarks 

```{r}
## Format data
# Return true/false where gene symbols in grch38 matches that in list of DEGs
idx <- grch38$symbol %in% rownames(ec.markers)
# Subset for rows in grch38 that matches DEGs 
ids <- grch38[idx,]

# The gene names can map to more than one Ensembl ID (some genes change ID over time), 
# so we need to remove duplicate IDs prior to assessing enriched GO terms
non_duplicates <- which(duplicated(ids$symbol) == FALSE)
ids <- ids[non_duplicates, ] 

# Add gene column in DEG list 
ec.markers["gene"] <- rownames(ec.markers)

# Merge DEGs with subset of table from grch38
res_ids <- inner_join(ec.markers, ids, by=c("gene" ="symbol")) 

# Create dataset containing significant genes 
sig_ids <- subset(res_ids, p_val_adj < 0.05)

# remove NAs
res_entrez <- subset(sig_ids, entrez != "NA")

# Remove duplicates 
res_entrez <- res_entrez[which(duplicated(res_entrez$entrez) == F), ]

# Extract name and fold-changes
foldchanges <- res_entrez$avg_log2FC

# Name each fold change with the corresponding Entrez ID
names(foldchanges) <- res_entrez$entrez

# Order in decreasing order
foldchanges <- sort(foldchanges, decreasing = TRUE)

## Load GMT file downloaded from MSigDB website 
h <- read.gmt("/Users/serachoi/Documents/Edinburgh/MScProject/Project/data/MSigDB_H")

# Hallmarks
msig_h_ec <- GSEA(foldchanges, 
               TERM2GENE=h,
               scoreType = "pos",
               minGSSize = 10,
               pvalueCutoff = 1,
               pAdjustMethod = "BH")

msig_H_result_ec <- data.frame(msig_h_ec)
msig_H_tidy_ec <- subset(msig_H_result_ec, subset = msig_H_result_ec$p.adjust<0.5)

#Barplot
ggplot(msig_H_tidy_ec, aes(reorder(Description, NES), NES)) + geom_col(aes(fill=p.adjust< 0.2)) + theme_minimal() + coord_flip() + theme(axis.text = element_text(size = 12)) + scale_fill_manual(values = c(mycoloursP[4],mycoloursP[5])) 
```


## GSEA with MSigDB Gene ontology

```{r}
## Load GMT file downloaded from MSigDB website 
c5 <- read.gmt("/Users/serachoi/Documents/Edinburgh/MScProject/Project/data/MSigDB_C5")

# Hallmarks
msig_c5_ec <- GSEA(foldchanges, 
               TERM2GENE=c5,
               scoreType = "pos",
               minGSSize = 10,
               pvalueCutoff = 1,
               pAdjustMethod = "BH")

msig_C5_result_ec <- data.frame(msig_c5_ec)

msig_C5_tidy_ec <- subset(msig_C5_result_ec, subset = msig_C5_result_ec$p.adjust<0.25)

# Remove GOBP and _ for prettier label for graph 
pathway <- rownames(msig_C5_tidy_ec)
pathway <- str_remove_all(pathway, "GOBP_")
pathway <- str_replace_all(pathway, "_", " ")
pathway <- tolower(pathway)
msig_C5_tidy_ec$Description <- pathway

#Barplot
pdf("/Users/serachoi/Documents/Edinburgh/MScProject/Project/outs/es-derived/GO_ec.pdf", width = 10, height = 8)
ggplot(msig_C5_tidy_ec[1:20,], aes(reorder(Description, NES), NES)) + geom_col(aes(fill=p.adjust< 0.2)) + theme_light() + coord_flip() + theme(axis.text = element_text(size = 18)) + scale_fill_manual(values = c(mycoloursP[5],mycoloursP[4])) 
dev.off()
```

# Visualisation

## Volcano plot 

```{r}
DefaultAssay(ec) <- "RNA"
Idents(ec) <- "orig.ident"
ec <- NormalizeData(ec)

all.markers <- FindMarkers(ec,
                           ident.1 = "McCracken",
                           ident.2 = "Luise",
                           only.pos = FALSE, 
                           min.pct = 0.25,
                           logfc.threshold = 0,
                           test.use = "MAST")

all.markers_sig <- subset(all.markers, subset = all.markers$p_val_adj < 0.0001)

# Format list of DEGs
all.markers$diff <- "NO"
all.markers$diff[all.markers$avg_log2FC > 2.5] <- "UP"
all.markers$diff[all.markers$avg_log2FC < -2.5] <- "DOWN"

pdf("/Users/serachoi/Documents/Edinburgh/MScProject/Project/outs/ES_volcano.pdf")
# Volcano plot 
ggplot(data = all.markers,aes(x=avg_log2FC, y=-log10(p_val_adj), col=diff)) + 
  geom_point() +
  theme_light() +
  geom_hline(yintercept = -log10(0.01), col="red") + 
  geom_vline(xintercept = c(-2.5,2.5), col="red") + 
  scale_color_manual(values=c(mycoloursP[2],mycoloursP[25],mycoloursP[1]))
dev.off()

```





