---
title: "03_Cracken_RC11_H9Day12"
author: "Sera Choi"
date: "14/07/2021"
output: html_document
---

```{r, message=FALSE, warning=FALSE}
# Load library
library(Seurat)
library(ggplot2)
library(dplyr)
library(here)
library("scales")
library(ggsci)
library(gtools)
library(future)

mypal <- pal_npg("nrc", alpha = 0.7)(10)
mypal2 <-pal_tron("legacy", alpha = 0.7)(7)
mypal3 <- pal_lancet("lanonc", alpha = 0.7)(9)
mypal4 <- pal_simpsons(palette = c("springfield"), alpha = 0.7)(16)
mypal5 <- pal_rickandmorty(palette = c("schwifty"), alpha = 0.7)(6)
mypal6 <- pal_futurama(palette = c("planetexpress"), alpha = 0.7)(5)
mypal7 <- pal_startrek(palette = c("uniform"), alpha = 0.7)(5)
mycoloursP<- c(mypal, mypal2, mypal3, mypal4, mypal5, mypal6, mypal7)
show_col(mycoloursP, labels =F)

# Set maximum size of the object 
options(future.globals.maxSize = 50000*1024^2)
```

# McCraken et al (GSE131736)

After analysing just H9 cell line Day 8, I've decided to expand the dataset to include RC11 day8 and H9 day 12. This section downloads the data from GEO, processes to subset ECs. The subset ECs will be converted to RDS format and saved in data/processed. 

## H9 Day 12

```{r}
counts <- Read10X_h5("/Users/serachoi/Documents/Edinburgh/MScProject/Project/data/Craken_H9_day12.h5", use.names = TRUE, unique.features = TRUE)
seur <- CreateSeuratObject(counts = counts, assay = "RNA", min.cells = 3, min.features = 200)
```

## Quality control

Using thresholds Craken et al used, cells with 

* cells expressing < 250 genes 
* cells having UMI count > 10,000
* cells with > 15% mito genes 

were removed. 

```{r}
seur[["percent.mt"]] <- PercentageFeatureSet(seur, pattern = "^MT-")
seur[["orig.ident"]] <- "McCracken_H9_Day12"
VlnPlot(seur, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3, pt.size =0)

# Apply QC threshold
seur <- subset(seur, subset = nCount_RNA > 250 & nFeature_RNA < 10000 & percent.mt < 15)

#Visualise
VlnPlot(seur, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3, pt.size =0)
plot1 <- FeatureScatter(seur, feature1 = "nCount_RNA", feature2 = "percent.mt")
plot2 <- FeatureScatter(seur, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")
plot1 + plot2
```

# Normalisation, scaling, reduction and visualisation

All done following McCracken et al's methods.

```{r}
# Normalisation
seur <- NormalizeData(seur, normalization.method = "LogNormalize", scale.factor = 10000)

# Feature Selection
seur <- FindVariableFeatures(seur, selection.method = "vst", nfeatures = 2000)

# Scaling 
all.genes <- rownames(seur)
seur <- ScaleData(seur, features = all.genes)

# Linear dimensional reduction
seur <- RunPCA(seur, features = VariableFeatures(object = seur), verbose = FALSE)
ElbowPlot(seur, ndims=50)

# Visualisation
seur <- RunUMAP(seur, dims = 1:20, verbose = FALSE)
DimPlot(seur, reduction = "umap")

# CLustering 
seur <- FindNeighbors(seur, dims = 1:20, verbose = FALSE)
seur <- FindClusters(seur, resolution = 0.01, verbose = FALSE)
DimPlot(seur, reduction = "umap")

saveRDS(seur, "/Users/serachoi/Documents/Edinburgh/MScProject/Project/data/McCracken_H9_day12.RDS")
```

## Cell-type characterisation

Using gene markers Craken et al used:

* pluripotent (POUF51, SOX2, NANOG)
* mesodermal (HAND1, T, MIXL1)
* mesenchymal (ACTA2, ALCAM, PDGFRB)
* endothelial (PECAM1, KDR, CD34, CDH5)
* epithelial (EPCAM, CLDN4, RAB25, WNT6)
* nephrogenic (PAX2, PAX8, CRYM)


```{r, fig.width=10}
# endothelial 
FeaturePlot(seur, features = c('PECAM1', 'KDR', 'CD34', 'CDH5'))
# Mesenchymal
FeaturePlot(seur, features = c('ACTA2', 'ALCAM', 'PDGFRB'))
```

## Subeset ECs

5,190 ECs

```{r}
Idents(seur) <- "RNA_snn_res.0.01"
es <- subset(seur, idents = 0)
saveRDS(es, "/Users/serachoi/Documents/Edinburgh/MScProject/Project/data/processed/McCracken_H9Day12_ec.RDS")
```

# RC11 Cell line 

```{r}
counts <- Read10X_h5("/Users/serachoi/Documents/Edinburgh/MScProject/Project/data/Cracken_RC11.h5", use.names = TRUE, unique.features = TRUE)
seur <- CreateSeuratObject(counts = counts, assay = "RNA", min.cells = 3, min.features = 200)
```

## Quality control

Using thresholds Craken et al used, cells with 

* cells expressing < 250 genes 
* cells having UMI count > 10,000
* cells with > 15% mito genes 

were removed. 

```{r}
seur[["percent.mt"]] <- PercentageFeatureSet(seur, pattern = "^MT-")
seur[["orig.ident"]] <- "McCracken_RC11"
VlnPlot(seur, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3, pt.size =0)

# Apply QC threshold
seur <- subset(seur, subset = nCount_RNA > 250 & nFeature_RNA < 10000 & percent.mt < 15)

#Visualise
VlnPlot(seur, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3, pt.size =0)
plot1 <- FeatureScatter(seur, feature1 = "nCount_RNA", feature2 = "percent.mt")
plot2 <- FeatureScatter(seur, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")
plot1 + plot2
```

# Normalisation, scaling, reduction and visualisation

All done following McCracken et al's methods.

```{r}
# Normalisation
seur <- NormalizeData(seur, normalization.method = "LogNormalize", scale.factor = 10000)

# Feature Selection
seur <- FindVariableFeatures(seur, selection.method = "vst", nfeatures = 2000)

# Scaling 
all.genes <- rownames(seur)
seur <- ScaleData(seur, features = all.genes)

# Linear dimensional reduction
seur <- RunPCA(seur, features = VariableFeatures(object = seur), verbose = FALSE)
ElbowPlot(seur, ndims=50)

# Visualisation
seur <- RunUMAP(seur, dims = 1:20, verbose = FALSE)
DimPlot(seur, reduction = "umap")

# CLustering 
seur <- FindNeighbors(seur, dims = 1:20, verbose = FALSE)
seur <- FindClusters(seur, resolution = 0.01, verbose = FALSE)
DimPlot(seur, reduction = "umap")

saveRDS(seur, "/Users/serachoi/Documents/Edinburgh/MScProject/Project/data/McCracken_RC11.RDS")
```

## Cell-type characterisation

Using gene markers Craken et al used:

* pluripotent (POUF51, SOX2, NANOG)
* mesodermal (HAND1, T, MIXL1)
* mesenchymal (ACTA2, ALCAM, PDGFRB)
* endothelial (PECAM1, KDR, CD34, CDH5)
* epithelial (EPCAM, CLDN4, RAB25, WNT6)
* nephrogenic (PAX2, PAX8, CRYM)


```{r, fig.width=10}
# endothelial 
FeaturePlot(seur, features = c('PECAM1', 'KDR', 'CD34', 'CDH5'))
# Mesenchymal
FeaturePlot(seur, features = c('ACTA2', 'ALCAM', 'PDGFRB'))
```

## Subeset ECs

2,307 ECs

```{r}
Idents(seur) <- "RNA_snn_res.0.01"
es <- subset(seur, idents = 1)
saveRDS(es, "/Users/serachoi/Documents/Edinburgh/MScProject/Project/data/processed/McCracken_RC11_ec.RDS")
```



