---
title: "11_SCT"
author: "Sera Choi"
date: "11/07/2021"
output: html_document
---

```{r, message=FALSE, warning=FALSE}
# Load library
library(Seurat)
library(ggplot2)
library(ggsci)
library(gtools)
library(dplyr)
library(here)
library("scales")
library(escape)
library(dittoSeq)
library(msigdbr)
library(fgsea)
library(tibble)
library(future)

mypal <- pal_npg("nrc", alpha = 0.7)(10)
mypal2 <-pal_tron("legacy", alpha = 0.7)(7)
mypal3 <- pal_lancet("lanonc", alpha = 0.7)(9)
mypal4 <- pal_simpsons(palette = c("springfield"), alpha = 0.7)(16)
mypal5 <- pal_rickandmorty(palette = c("schwifty"), alpha = 0.7)(6)
mypal6 <- pal_futurama(palette = c("planetexpress"), alpha = 0.7)(5)
mypal7 <- pal_startrek(palette = c("uniform"), alpha = 0.7)(5)
mycoloursP<- c(mypal, mypal2, mypal3, mypal4, mypal5, mypal6, mypal7)
show_col(mycoloursP, labels =F)

# Set maximum size of the object 
options(future.global.maxSize = 100000*1024^2)
```

# Age analysis with SCTransform integrated data

Previous age analysis using integrated data (Luise, Yang, Lau) with CCA integration seemed strange. Here, I use SCT integration and normalisation.

## Merge Luise, Yang and Lau

Here I process ECs from Lau et al using SCT normalisation and integration 

```{r}
luise <- readRDS("/Users/serachoi/Documents/Edinburgh/MScProject/Project/data/processed/Luise_ec.RDS")
lau <- readRDS("/Users/serachoi/Documents/Edinburgh/MScProject/Project/data/processed/Lau_ec.RDS")
yang <- readRDS("/Users/serachoi/Documents/Edinburgh/MScProject/Project/data/processed/yang_capillary_ec.RDS")

# Organise metadata so sample IDs all under column BBN
lau[["BBN"]] <- lau[["orig.ident"]]
yang[["BBN"]] <- yang[["orig.ident"]]

# Set orig.ident
luise[["orig.ident"]] <- "Luise"
lau[["orig.ident"]] <- "Lau"
yang[["orig.ident"]] <- "Yang"

# Merge data
comb <- merge(x = luise, y = c(lau, yang), add.cell.ids = c("Luise", "Lau", "Yang"))
```

## SCT normalisation and integration

```{r}
# Set up Seurat object list, run SCTransform oneach object separately
ec.list <- SplitObject(comb, split.by = "orig.ident")
for (i in 1:length(ec.list)) {
    ec.list[[i]] <- SCTransform(ec.list[[i]], verbose = TRUE)
}

# Select features for downstream integration
ec.features <- SelectIntegrationFeatures(object.list = ec.list, nfeatures = 2000)
# Ensures that all necessary Pearson residuals have been calculated 
ec.list <- PrepSCTIntegration(object.list = ec.list, anchor.features = ec.features, verbose = TRUE)

# identify anchors and integrate datasets 
ec.anchors <- FindIntegrationAnchors(object.list = ec.list, normalization.method = "SCT", anchor.features = ec.features, verbose = TRUE)
ec.integrated <- IntegrateData(anchorset = ec.anchors, normalization.method = "SCT", verbose = TRUE)
```

## Clustering and visualisation 

```{r}
ec.integrated <- RunPCA(ec.integrated, verbose = FALSE)
ElbowPlot(ec.integrated)

ec.integrated <- RunUMAP(ec.integrated, dims = 1:20)
DimPlot(ec.integrated, reduction = "umap", group.by = "BBN", cols=mycoloursP[1:25])
```

## Downsampling Yang

```{r}
Idents(yang) <- "BBN"
reduced_yang <- subset(yang, downsample = 100)

# Merge data
comb <- merge(x = luise, y = c(lau, reduced_yang), add.cell.ids = c("Luise", "Lau", "Yang"))
```

## SCT normalisation and integration 

```{r}
# Set up Seurat object list, run SCTransform oneach object separately
ec.list <- SplitObject(comb, split.by = "orig.ident")
for (i in 1:length(ec.list)) {
    ec.list[[i]] <- SCTransform(ec.list[[i]], verbose = FALSE)
}

# Select features for downstream integration
ec.features <- SelectIntegrationFeatures(object.list = ec.list, nfeatures = 2000)
# Ensures that all necessary Pearson residuals have been calculated 
ec.list <- PrepSCTIntegration(object.list = ec.list, anchor.features = ec.features, verbose = FALSE)

# identify anchors and integrate datasets 
ec.anchors <- FindIntegrationAnchors(object.list = ec.list, normalization.method = "SCT", anchor.features = ec.features, verbose = FALSE)
ec.integrated <- IntegrateData(anchorset = ec.anchors, normalization.method = "SCT", verbose = FALSE)

# Normalise RNA data for visual purposes 
DefaultAssay(ec.integrated) <- "RNA"
ec.integrated <- NormalizeData(ec.integrated, verbose = FALSE)
```

## Quality control

```{r}
ec.integrated[["percent.mt"]] <- PercentageFeatureSet(ec.integrated, pattern = "^MT-", assay = "RNA")
VlnPlot(ec.integrated, features = "nFeature_RNA", group.by = "orig.ident", pt.size=0, cols = mycoloursP[1:25])
VlnPlot(ec.integrated, features = "nCount_RNA", group.by = "orig.ident", pt.size=0, cols = mycoloursP[1:25])
VlnPlot(ec.integrated, features = "percent.mt", group.by = "orig.ident", pt.size=0,cols = mycoloursP[1:25])

FeatureScatter(ec.integrated, feature1 = "nCount_RNA", feature2 = "percent.mt", group.by = "orig.ident",cols = mycoloursP[1:25])
FeatureScatter(ec.integrated, feature1 = "nCount_RNA", feature2 = "nFeature_RNA", group.by = "orig.ident",cols = mycoloursP[1:25])
```

## Clustering and visualisation 

```{r}
DefaultAssay(ec.integrated) <- "integrated"
ec.integrated <- RunPCA(ec.integrated, verbose = FALSE)
ElbowPlot(ec.integrated)
ec.integrated <- RunUMAP(ec.integrated, dims = 1:20)
```

```{r}

# Classify age group
mod_metadata <- ec.integrated@meta.data  %>% mutate(AgeGroup = case_when(Age < 65 ~ 'Young',
                                                                        Age > 65 ~ 'Old'))
                                    
ec.integrated[["AgeGroup"]] <- mod_metadata$AgeGroup

# ribo genes 
rpl <- PercentageFeatureSet(ec.integrated, pattern = "^RPL", assay = "RNA")
rps <- PercentageFeatureSet(ec.integrated, pattern = "^RPS", assay = "RNA")
ec.integrated[["percent.ribo"]] <- rpl+rps

# Individuals 
DimPlot(ec.integrated, reduction = "umap", group.by = "BBN",cols = mycoloursP[1:25])
# Dataset
DimPlot(ec.integrated, reduction = "umap", group.by = "orig.ident",cols = mycoloursP[1:25])
# Agegroup
DimPlot(ec.integrated, reduction = "umap", group.by = "AgeGroup",cols = mycoloursP[1:25])
# Tissue 
mod_metadata <- ec.integrated@meta.data  %>% mutate(Tissue = case_when(orig.ident == "Luise" ~ 'BA4',
                                                            orig.ident == "Lau" ~ "Prefrontal cortex",
                                                            orig.ident == "Yang" ~ "Superial frontal cortex"))
                                    
ec.integrated[["Tissue"]] <- mod_metadata$Tissue
DimPlot(ec.integrated, reduction = "umap", group.by = "Tissue",cols = mycoloursP[1:25])

# number of unique genes 
FeaturePlot(ec.integrated, features = "nFeature_RNA")
FeaturePlot(ec.integrated, features = "nCount_RNA")
FeaturePlot(ec.integrated, features = "percent.mt")
FeaturePlot(ec.integrated, features = "percent.ribo")

# Save RDS
saveRDS(ec.integrated, "/Users/serachoi/Documents/Edinburgh/MScProject/Project/data/SCT_ec")
```





















