---
title: "12_SCT_age"
author: "Sera Choi"
date: "11/07/2021"
output: html_document
---
```{r, message=FALSE, warning=FALSE}
# Load library
library(Seurat)
library(ggplot2)
library(ggsci)
library(gtools)
library(dplyr)
library(here)
library("scales")
library(escape)
library(dittoSeq)
library(tibble)
library(future)
library(clusterProfiler)
library(org.Hs.eg.db)
library(DOSE)
library(pathview)
library(clusterProfiler)
library(AnnotationHub)
library(ensembldb)
library(tidyverse)
library(annotables)
library(enrichplot)

mypal <- pal_npg("nrc", alpha = 0.7)(10)
mypal2 <-pal_tron("legacy", alpha = 0.7)(7)
mypal3 <- pal_lancet("lanonc", alpha = 0.7)(9)
mypal4 <- pal_simpsons(palette = c("springfield"), alpha = 0.7)(16)
mypal5 <- pal_rickandmorty(palette = c("schwifty"), alpha = 0.7)(6)
mypal6 <- pal_futurama(palette = c("planetexpress"), alpha = 0.7)(5)
mypal7 <- pal_startrek(palette = c("uniform"), alpha = 0.7)(5)
mycoloursP<- c(mypal, mypal2, mypal3, mypal4, mypal5, mypal6, mypal7)
show_col(mycoloursP, labels =F)

# Set maximum size of the object 
options(future.global.maxSize = 100000*1024^2)
```


# Age analysis of SCT integrated data

Rather than CCA, better, more advanced integration method SCTransform was used to integrate Luise, Lau and downsampled Yang data. Age analysis will be performed here. 

```{r}
ec <- readRDS("/Users/serachoi/Documents/Edinburgh/MScProject/Project/data/SCT_ec")
```

## Enhanced genes in Aged 

Previously using only Luise's dataset, only 8 gene were identified to be enhanced in aged cells. However, with increased cell depth, many more cells are identified to be enhanced. Overall 208 DEGs were identified. Of these, 50 were statistically significant (<0.05). I reduced the list furtherby setting thresholds:

* adjusted p-value < 0.05
* pct.1 > 25%
* pct.2 < 25%
* logFC > 0.25

Giving, overall 30 genes that are enhanced in aged ECs. 

* MKL2 was removed from the list because MKL2 and MRTFB are same the genes with different gene symbols. It appears that MKL2 is expressed by all cells across different datasets but in Lau and Yang, genes are labelled MKL2, while in Luise, gene is labelled MRTFB. Since Lau and Yang mostly represent Aged cells, MKL2 appears to be expressed more in Aged when in fact not. 

Therefore, 29 genes were enhanced in aged ECs. 

```{r, fig.height=13, fig.width=15}
DefaultAssay(ec) <- "RNA"
Idents(ec) <- "AgeGroup"

old.markers <- FindMarkers(ec,
                           ident.1 = "Old",
                           ident.2 = "Young",
                           only.pos = TRUE, 
                           min.pct = 0.25, 
                           logfc.threshold = 0.25,
                           test.use = "MAST")
old.markers

old.markers_sig <- subset(old.markers, old.markers$p_val_adj < 0.05 )
old.markers_stringent <- subset(old.markers, subset =(old.markers$p_val_adj < 0.05 & old.markers$pct.1 > 0.25 & old.markers$pct.2 < 0.25))
old.markers_stringent <- old.markers_stringent[c(-1),] # Remove MKL2

# Output 
old.markers_stringent
VlnPlot(ec, assay = "RNA", features = rownames(old.markers_stringent)[1:12], group.by = "AgeGroup", pt.size = 0)
```

## GSEA with MSigDB hallmarks and ontology terms 

GSEA was performed using 50 significant up-regulated genes in Aged (old.markers_sig).

* H = hallmark gene sets 
* c5 = gene sets derived from the GO biological process ontology

```{r}
## Format data
# Return true/false where gene symbols in grch38 matches that in list of DEGs
idx <- grch38$symbol %in% rownames(old.markers)
# Subset for rows in grch38 that matches DEGs 
ids <- grch38[idx,]

# The gene names can map to more than one Ensembl ID (some genes change ID over time), 
# so we need to remove duplicate IDs prior to assessing enriched GO terms
non_duplicates <- which(duplicated(ids$symbol) == FALSE)
ids <- ids[non_duplicates, ] 

# Add gene column in DEG list 
old.markers["gene"] <- rownames(old.markers)

# Merge DEGs with subset of table from grch38
res_ids <- inner_join(old.markers, ids, by=c("gene" ="symbol")) 

# Create dataset containing significant genes 
sig_ids <- subset(res_ids, p_val_adj < 0.05)

# remove NAs
res_entrez <- subset(sig_ids, entrez != "NA")

# Remove duplicates 
res_entrez <- res_entrez[which(duplicated(res_entrez$entrez) == F), ]

# Extract name and fold-changes
foldchanges <- res_entrez$avg_log2FC

# Name each fold change with the corresponding Entrez ID
names(foldchanges) <- res_entrez$entrez

# Order in decreasing order
foldchanges <- sort(foldchanges, decreasing = TRUE)

## Load GMT file downloaded from MSigDB website 
h <- read.gmt("/Users/serachoi/Documents/Edinburgh/MScProject/Project/data/MSigDB_H")

# Hallmarks
msig_h <- GSEA(foldchanges, 
               TERM2GENE=h,
               scoreType = "pos",
               minGSSize = 1,
               pvalueCutoff = 1,
               pAdjustMethod = "BH"
               )

msig_H_result <- data.frame(msig_h)
msig_H_result <- msig_H_result[-c(16),] # remove p53 pathway
msig_H_tidy <- subset(msig_H_result, subset = msig_H_result$pvalue<0.5)

#Barplot
ggplot(msig_H_result, aes(reorder(Description, NES), NES)) + geom_col(aes(fill=p.adjust< 0.2)) + theme_minimal() + coord_flip() + theme(axis.title = element_text(size = 12))
```

## GSEA with MSigDB gene ontology

```{r, fig.width=15, fig.height=5}
# Gene ontology
c5 <- read.gmt("/Users/serachoi/Documents/Edinburgh/MScProject/Project/data/MSigDB_C5")

msig_c5 <- GSEA(foldchanges, 
               TERM2GENE=c5,
               scoreType = "pos",
               minGSSize = 2,
               pvalueCutoff = 1,
               pAdjustMethod = "BH"
               )

msig_C5_result <- data.frame(msig_c5)
msig_C5_tidy <- subset(msig_C5_result, subset = msig_C5_result$pvalue<0.05)

#Barplot
ggplot(msig_C5_tidy, aes(str_wrap(Description,width=1), NES)) + geom_col(aes(fill=p.adjust <0.2)) + theme_minimal() + coord_flip()
```

# Enhanced genes in Young 

With MAST, 315 enhanced genes found in young cells. Of these, 62 are significant under adjusted p-val < 0.05. After applying more stringent thresholds same as aged cells, 22 are significant. 

Without any thresholds, many genes are mito related, perhaps cells are stressed.
After adjusted p-val <0.05 is applied, mito genes still remain.
It's only after the stringent thresholds are applied, most of mito genes disappear. But now, ribosomal genes appear. 


```{r, fig.height=13, fig.width=15}
Idents(ec) <- "AgeGroup"

young.markers <- FindMarkers(ec,
                           assay = "RNA",
                           ident.1 = "Young",
                           ident.2 = "Old",
                           only.pos = TRUE, 
                           min.pct = 0.25, 
                           logfc.threshold = 0.25,
                           test.use = "MAST")
young.markers

young.markers_sig <- subset(young.markers, young.markers$p_val_adj < 0.05 )
young.markers_sig

young.markers_stringent <- subset(young.markers, subset =(young.markers$p_val_adj < 0.05 & 
                                                            young.markers$pct.1 > 0.25 & 
                                                            young.markers$pct.2 < 0.25))
young.markers_stringent <- young.markers_stringent[c(-1),] # remove MKL2

VlnPlot(ec, assay = "RNA", features = rownames(young.markers_stringent)[1:12], group.by = "AgeGroup", pt.size = 0)
```

### Checking Ribo genes 

It is quite strange to have so many ribo genes reported as enhanced in young. It could be that the cells are too similar and the statistics picks up only the most abundant genes just because they are most easy to detect. Therefore, here I checked whether the % of total RPL/RPS genes is different compared to non-RPS/RPL genes. Perhaps there aren't many genes that are enahcned in young when compared to old. Since almost all (Expet 1) indiciduals in young group are from Luise's data and luise's dataset contains high percentage of ribosomal genes, perhaps it is finding genes that are different between datasets (Luise vs Lau+Yang). So, apart from first two, the rest are disregarded.  

```{r}
VlnPlot(ec, features = "percent.ribo", group.by = "orig.ident")
VlnPlot(ec, features = "percent.ribo", group.by = "AgeGroup")
VlnPlot(ec, features = c("RPL37","RPS21"), group.by = "AgeGroup")

# How many RPS/RPL genes 
library(stringr)
DefaultAssay(ec) <- "RNA"
rpl_genes <- str_detect(rownames(ec), pattern = "^RPL")
rps_genes <- str_detect(rownames(ec), pattern = "^RPS")
ribo_percent <- (sum(rpl_genes) + sum(rps_genes))*100/length(rownames(ec))
ribo_percent
```

# GSEA with MSigDB Hallmarks 


```{r}
## Format data
# Return true/false where gene symbols in grch38 matches that in list of DEGs
idx <- grch38$symbol %in% rownames(young.markers)
# Subset for rows in grch38 that matches DEGs 
ids <- grch38[idx,]

# The gene names can map to more than one Ensembl ID (some genes change ID over time), 
# so we need to remove duplicate IDs prior to assessing enriched GO terms
non_duplicates <- which(duplicated(ids$symbol) == FALSE)
ids <- ids[non_duplicates, ] 

# Add gene column in DEG list 
young.markers["gene"] <- rownames(young.markers)

# Merge DEGs with subset of table from grch38
res_ids <- inner_join(young.markers, ids, by=c("gene" ="symbol")) 

# Create dataset containing significant genes 
sig_ids <- subset(res_ids, p_val_adj < 0.05)

# remove NAs
res_entrez <- subset(sig_ids, entrez != "NA")

# Remove duplicates 
res_entrez <- res_entrez[which(duplicated(res_entrez$entrez) == F), ]

# Extract name and fold-changes
foldchanges <- res_entrez$avg_log2FC

# Name each fold change with the corresponding Entrez ID
names(foldchanges) <- res_entrez$entrez

# Order in decreasing order
foldchanges <- sort(foldchanges, decreasing = TRUE)

## Load GMT file downloaded from MSigDB website 
h <- read.gmt("/Users/serachoi/Documents/Edinburgh/MScProject/Project/data/MSigDB_H")

# Hallmarks
msig_h_young <- GSEA(foldchanges, 
               TERM2GENE=h,
               scoreType = "pos",
               minGSSize = 1,
               pvalueCutoff = 1,
               pAdjustMethod = "BH"
               )

msig_H_result_young <- data.frame(msig_h_young)
msig_H_tidy_young <- subset(msig_H_result_young, subset = msig_H_result_young$pvalue<0.7)

#Barplot
ggplot(msig_H_tidy_young, aes(reorder(Description, NES), NES)) + geom_col(aes(fill=p.adjust< 0.2)) + theme_minimal() + coord_flip() + theme(axis.title = element_text(size = 12))
```


## GSEA with MSigDB Gene Onotology

```{r, fig.width=15}
## Load GMT file downloaded from MSigDB website 

# Hallmarks
msig_c5_young <- GSEA(foldchanges, 
               TERM2GENE=c5,
               scoreType = "pos",
               minGSSize = 2,
               pvalueCutoff = 1,
               pAdjustMethod = "BH"
               )

msig_C5_result_young <- data.frame(msig_c5_young)
msig_C5_tidy_young <- subset(msig_C5_result_young, subset = msig_C5_result_young$p.adjust<0.1)

#Barplot
ggplot(msig_C5_tidy_young[1:20,], aes(reorder(Description, NES), NES)) + geom_col(aes(fill=p.adjust< 0.2)) + theme_minimal() + coord_flip() + theme(axis.title = element_text(size = 12))
```


# Visualisations


## Bar Plot showing log2FC 

Can have ribosomal genes removed or retained. 

```{r, fig.width=20, fig.height=6}
DefaultAssay(ec) <- "RNA"
Idents(ec) <- "AgeGroup"

all.markers <- FindMarkers(ec,
                           assay = "RNA",
                           ident.1 = "Old",
                           ident.2 = "Young",
                           only.pos = FALSE, 
                           min.pct = 0.25, 
                           logfc.threshold = 0.25,
                           test.use = "MAST")

# Obtain top 10 enhanced genes in young cells
all.markers_young <- subset(all.markers, subset= all.markers$avg_log2FC<0)
all.markers_young <- subset(all.markers_young, subset = (all.markers_young$p_val_adj < 0.05 & all.markers_young$pct.1<0.25& all.markers_young$pct.2>0.25))
all.markers_young <- all.markers_young[c(-1),] # Remove MKL2


# Combine enhanced genes in old and young 
all.markers <- rbind(old.markers_stringent[1:20,], all.markers_young[1:20,])
all.markers$gene <- rownames(all.markers)
all.markers[order(all.markers$avg_log2FC, decreasing = TRUE),]
all.markers$log10_FDR <- -log10(all.markers$p_val_adj)

#pdf("/Users/serachoi/Documents/Edinburgh/MScProject/Project/outs/Age_barplot.pdf", width = 15, height = 5)

# Barplot
ggplot(data = all.markers, aes(x = reorder(gene, -avg_log2FC), y = avg_log2FC, fill = log10_FDR)) + geom_bar(stat="identity")  + theme_light()+ theme(axis.text = element_text(size=13), axis.text.x = element_text(angle = 45, vjust = 0.5, hjust=0.4), legend.text = element_text(size = 12), legend.title = element_text(size = 13), text = element_text(size = 15)) + scale_fill_continuous(type="viridis")
#dev.off()

```

## DotPlot of enhanced genes in Aged and Young

This plot shows top 10 enhanced genes in young and old.

```{r, fig.height=5, fig.width=12}

markers <- rownames(all.markers)
pdf("/Users/serachoi/Documents/Edinburgh/MScProject/Project/outs/Age_dotplot.pdf", height = 5, width = 12)
dittoDotPlot(object = ec, 
             vars = markers, 
             group.by = "AgeGroup", 
             scale = F,
             min.color = "navyblue",
             max.color = "orangered",
             size = 10,
             assay = "RNA") + theme_light() + theme(axis.text = element_text(size=13), legend.text = element_text(size = 12), axis.text.x = element_text(angle = 45, vjust = 0.5, hjust=0.4)) 
dev.off()
```





