---
title: "04_annotation_Luise"
author: "Sera Choi"
date: "24/05/2021"
output: html_document
---

```{r, message=FALSE, warning=FALSE}
# Load library
library(Seurat)
library(ggplot2)
library(dplyr)
library(SingleCellExperiment)
library(scater) 
library(here)
library("scales")
library(ggsci)
library(gtools)

# Load data
ecpc <- readRDS(here("/Users/serachoi/Documents/Edinburgh/MScProject/Project/data/processed/ecpc03.RDS")) 
```

# Annotation at resolution = 0.2 
In this section, three clusters at resolution 0.2 will be annotated by identifying genes that drive the separation between the clusters. 

```{r, message=FALSE, warning=FALSE}
ecpc_0.2 <- FindClusters(ecpc, resolution = 0.2 )
ecpc_0.2 <- RunUMAP(ecpc_0.2, dims = 1:25)
DimPlot(ecpc_0.2, reduction = "umap", label=TRUE, label.size = 8) + NoLegend()
```

# Gene Marker Identification

Seurat uses function FindAllMarkers() to identify positive and negative gene markers of one cluster, compared to all other cells. So, if a gene in cluster 0 has avg.log2FC of 2.0 with highly significant adjusted p-value, then, cells in cluster 0 express the gene on average 2-fold more than cells in cluster 1 and 2.    

* min.pct : if min.pct = 0.25, genes that are expressed in 25% of cells in either of the two populations are only tested. This reduces computational cost as irrelavant genes are not tested.
* logfc.threshold : only those genes with average log X-fold difference between two groups of cells are tested.  
* only positive gene markers are returned. 
* By default, Wilcoxon Rank Sum test is used but here, I chose to use MAST because literatures have identified MAST to be best performing single-cell DE testing method. 

```{r, message=FALSE}
# Set cell identity class to clustering resolution 0.2
Idents(ecpc_0.2) <- "RNA_snn_res.0.2"

# Finds gene markers in each cluster compared to all other cells 
ecpc_0.2.markers <- FindAllMarkers(ecpc_0.2, 
                               only.pos = TRUE, 
                               min.pct = 0.25, 
                               logfc.threshold = 0.25,
                               test.use = "MAST")
```

The results below show 10 gene markers for each cluster ordered with highest log2FC: 

* unadjusted p-value 
* log2 fold-change of average expression between two groups. Positive value indicates that cells in first group express the feature more 
* pct1: percentage of cells that express the feature in group 1
* pct2: percentage of cells that express the feature in group 2
* adjusted p-value after Bonferroni correction 

Since only positive gene markers are reported here, pct.1 > pct.2

```{r, message=FALSE}
top10_markers_0.2 <- ecpc_0.2.markers %>% group_by(cluster) %>% top_n(n = 10, wt = avg_log2FC)
top10_markers_0.2
```

# Feature plots with known markers  

Here, I created feature plots for clusters 0, 1 and 2 using known markers (EC, pericytes and SMCs) and markers identified in DE analysis above (top 10 with highest log2FC). Gene markers for ECs and pericytes were found in literatures (Luise and 'Reconstruction of the human blood–brain barrier in vitro reveals a pathogenic mechanism of APOE4 in pericytes', 2020) :

* Endothelial:CLDN5, VWF, FLT1, ANO2, ST6GALNAC3, ELOVL7 and ABCB1
* Pericytes: PDGFRB, NOTCH3, PRKG1, LAMA2, GPC5, PDE7B, and DLC1
* SMCs: MYH11, TAGLN, ACTA2, LPP and MYL9

## Endothelial 

* No distinct gene marker for this cluster
* Scattered but expressed slightly more in cells in cluster 0

```{r, fig.width= 15, fig.height= 13}
FeaturePlot(ecpc, features= c("CLDN5", 
                               "VWF", 
                               "FLT1", 
                               "ANO2", 
                               "ST6GALNAC3", 
                               "ELOVL7", 
                               "ABCB1"))
```

## Pericyte

* More concentrated than EC markers 
* Highly expressed in cells in cluster 1

```{r, fig.width= 15, fig.height= 13}
FeaturePlot(ecpc, features= c("PDGFRB", 
                              "NOTCH3", 
                              "PRKG1", 
                              "LAMA2", 
                              "GPC5", 
                              "PDE7B",
                              "DLC1"))
```

## Smooth Muscle cell (vascular cells)

* cells in sub-cluster of cluster 2 appears to express SMC related genes 
* Maybe increase resolution to capture this? 

```{r, fig.width= 15, fig.height= 13}
FeaturePlot(ecpc, features= c("MYH11", 
                              "TAGLN", 
                              "ACTA2", 
                              "LPP", 
                              "MYL9"))
```

# Feature plots with identified markers 

Here I created feature plots for cluster 0, 1 and 2 using 10 identified gene markers from DE analysis (top 10 logFC). I am mostly interested in identifying cluster 2 because cluster 2 was neither ECs, pericytes or SMCs. 

## Cluster 0 

* No distinct gene marker for this cluster 
* Known gene markers for ECs seen (ABCB1 and THSD4): further confirms that cluster 0 is ECs

```{r}
# This function takes result from DE analysis and creates feature plots 
DEG_feature_plot_func <- function(seur_object,
                                  markers){
  
  # Converts data.frame into string of DEGs
  gene.markers <- as.character(as.matrix(markers))
  
  # Create feature plot using makers 
  feat_plot <- FeaturePlot(seur_object, 
                           features = c(gene.markers))
  print(feat_plot)
}
```


```{r, fig.width= 15, fig.height= 13}
# Cluster 0 annotation with 10 identified markers 
clust0_markers_0.2 <- top10_markers_0.2[1:10, 'gene']
DEG_feature_plot_func(ecpc_0.2, clust0_markers_0.2)
```

## Cluster 1 gene marker 

* Known pericyte markers seen (PRKG1, GPC5, DLC1)
* genes related to SMCs also seen (TAGLN, ACRA2) 

```{r, fig.width= 15, fig.height= 13}
# Cluster 1 top 10 gene markers 
clust1_markers_0.2 <- top10_markers_0.2[11:20, 'gene']
DEG_feature_plot_func(ecpc_0.2, clust1_markers_0.2)
```

## Cluster 2 gene marker 

* CXCL8 and SELE almost exclusively in cluster 2
* CXCL8: Interleukin expressed by ECs. 
* SELE: E-selectin expressed by ECs. 

```{r, fig.width= 15, fig.height= 13}
# Cluster 2 top 10 gene markers 
clust2_markers_0.2 <- top10_markers_0.2[21:30, 'gene']
DEG_feature_plot_func(ecpc_0.2, clust2_markers_0.2)
```

# Number of cells per cluster 

Significantly more cells from young samples in cluster 2. When we want to investigate whether the cells are clustering by technical reasons (age, gender, sameples), we look at sample level. There is no cause of concern since we have so little cells and those cells in clsuter 2 are interesting, continue. 

```{r}
# Number of cells per cluster at resolution 0.2
nCellsPerCluster <- table(Idents(ecpc_0.2))
nCellsPerCluster
```

```{r}
# Number of cells per cluster by age group
nCellsAgePerCluster <- table(Idents(ecpc_0.2), ecpc_0.2$AgeGroup)
nCellsAgePerCluster

nCellsSamplePerCluster <- table(Idents(ecpc_0.2), ecpc_0.2$BBN)
nCellsSamplePerCluster
```

## Visualise by Age Group, Gender and Samples

Two samples 1.30841 and BBN14396 contribute most to cluster 2- but again, not an issue.  

```{r}
DimPlot(ecpc, reduction='umap',group.by = "AgeGroup")
DimPlot(ecpc, reduction='umap',group.by = "gender")
DimPlot(ecpc, reduction='umap',group.by = "BBN")
```

# Resolution = 0.25 and 0.3

Resolution 0.25 identifies 4th cluster, which are the SMCs. At resolution 0.3, another sub-population of ECs appear in EC population, which seem quite interesting. So, I follow onto identify those sub-clusters. 

```{r, fig.width=10, fig.height=5, warning=FALSE, message=FALSE }

ecpc_0.25 <- FindClusters(ecpc, resolution = 0.25)
ecpc_0.25 <- RunUMAP(ecpc_0.25, dims = 1:25)

ecpc_0.3 <- FindClusters(ecpc, resolution = 0.3)
ecpc_0.3 <- RunUMAP(ecpc_0.3, dims = 1:25)

ecpc_0.25_plot <- DimPlot(ecpc_0.25, 
                          reduction = "umap", 
                          label=TRUE, 
                          label.size = 8) + NoLegend()

ecpc_0.3_plot <- DimPlot(ecpc_0.3, 
                         reduction = "umap", 
                         label=TRUE, 
                         label.size = 8) + NoLegend()

multiplot(ecpc_0.25_plot, ecpc_0.3_plot, cols=2)
```

# Annotation at resolution = 0.3 

Here, I attempt to identify the two sub-clusters in endothelial population. It may be that those sub-clusters belong to different vessel types- arteriolar and venular ECs. Therefore, first I tried to identify using known arteriolar and venular gene markers. However, not many human gene markers are known (many mice and zebra fish). Few were taken from Arterial versus venous endothelial cells (2014).

* Arterial ECs: HEY2, GJA4, 
* Venous ECs: NR2F2, EMCN 

Another study that compared coronary artery and aortic ECs against umbilical vein ECs (Identification of endothelial cell genes by combined database mining and microarray analysis, 2013) has noted that:

* Arterial ECs have over-expression of thrombomodulin and aldo-keto reductase family member 3
* Venular ECs have over-expression of matrix metalloprotease 10, integrins and adhesion proteins. 

It would also be interesting to know whether the expression profile of pericytes also differ with vessel types at resolution 0.5 where pericyte population also divides into two. 

## Feature plot with known markers 

Arterivenous differentiation occurs during embryonic development and differentiation of angioblasts leads to formation of network of arteries, veins and capillaries. Notch signalling pathway appears to have importance in deciding the cell fate. Several genes in nothch pathway are up-regulated in arterial ECs than venous ECs. Hey2 is a downstream target of Notch signalling exclusively expressed in arterial ECs, while COUP-TFII (NR2F2) is exclusively expressed in venous ECs (Impact of Hey2 and COUP-TFII on genes involved in arteriovenous differentiation in primary human arterial and venous endothelial cells, 2013). 

### Artery 

No distinct clusters express arterial gene markers. Interestingly, pericytes are expressing more arterial gene markers than cells in EC population. Expression of HEY2 is increased with Notch signal pathway. Maybe HEY2 is required during angiogenesis but for fully differenciated arterial ECs, HEY2 isn't expressed. Potentially, the same for GJA4. 

```{r,fig.width=10, fig.height=5}
Idents(ecpc_0.3) <- "RNA_snn_res.0.3"
FeaturePlot(ecpc_0.3, features = c("HEY2", "GJA4"))
```

#### Notch3 communication between ECs and pericytes 

It could be that pericytes are expressing HEY2 because of communication between ECs and pericytes. DLL4 is a Notch ligand expressed by ECs. DLL4 binds to NOTCH3 receptors on pericytes, which increases downstream genes, HEY2 and HEYL. Enhanced Notch3 signalling in pericytes enhanced adhesion strength of pericytes to ECs and decreased migratory ability.  This study was down in human brain pericytes. (Cerebral Cavernous Malformation-1 Protein Controls DLL4-Notch3 Signaling Between the Endothelium and Pericytes, 2015).

So, perhaps HEY2 expression isn't sustained after fully differentiating and the expression of HEY2 in pericyte is to do with communication between pericytes and ECs. 

```{r,fig.width=10, fig.height=6}
FeaturePlot(ecpc, feature= c("DLL4",
                             "NOTCH3",
                             "HEY2",
                             "HEYL"))
```


### Vein

NR2F2 (COUP-TFII) is a well-known venous EC marker during angiogenesis. However, again it is expressed in pericyte population more than EC population. This may be because NR2F2 level is no longer sustained after differentiation. I think NR2F2 is involved in communication with ECs. 

```{r,fig.width=10, fig.height=5}
FeaturePlot(ecpc_0.3, features = c('NR2F2', 'EMCN'))
```

#### Communication between Pericyte and ECs

NR2F2 is expressed by pericytes, which increases expression of ANGPT1 (paracrine ligand specific for endothelium-specific tyrosine kinase receptor TEK). (The role of the orphan nuclear receptor COUP-TFII in tumorigenesis, 2015).  

```{r,fig.width=10, fig.height=6}
FeaturePlot(ecpc, feature=c("NR2F2",
                            "ANGPT1",
                            "TEK"))
```

# Pairwise comparison of cluster 1 and 2

Cluster annotation using known gene markers was unsuccessful, so I performed pair-wise comparison of cluster 1 and 2 to find DEGs. 

## Cluster 1

```{r, fig.width= 15, fig.height= 13, message=FALSE}
Idents(ecpc_0.3) <- "RNA_snn_res.0.3"

clust1_markers_0.3 <- FindMarkers(ecpc_0.3, 
                                  ident.1 = 1, 
                                  ident.2 = 2, 
                                  only.pos = TRUE,
                                  min.pct = 0.25,
                                  logfc.threshold = 0.25,
                                  test.use = "MAST")

# Order DEGs by average log2FC
clust1_markers_0.3 <- clust1_markers_0.3[order(clust1_markers_0.3$avg_log2FC),]

# Top 10 DEGs
top10_clust1_markers_0.3 <- rownames(clust1_markers_0.3[1:10,])

# Create feature plot using top 10 DEGs
DEG_feature_plot_func(ecpc_0.3, top10_clust1_markers_0.3)
```

## Cluster 2

```{r, fig.width= 15, fig.height= 13, message=FALSE}

clust2_markers_0.3 <- FindMarkers(ecpc_0.3, 
                                  ident.1 = 2, 
                                  ident.2 = 1, 
                                  only.pos = TRUE,
                                  min.pct = 0.25,
                                  logfc.threshold = 0.25,
                                  test.use = "MAST")

# Order DEGs by average log2FC
clust2_markers_0.3 <- clust2_markers_0.3[order(clust2_markers_0.3$avg_log2FC),]

# Top 10 DEGs
top10_clust2_markers_0.3 <- rownames(clust2_markers_0.3[1:10,])

# Create feature plot using top 10 DEGs
DEG_feature_plot_func(ecpc_0.3, top10_clust2_markers_0.3)

```

# Feature Plot with UMI and gene counts 

No clear genes that differentiate cluster 1 and 2 of EC population could be identified. To check whether the cells are clustering due to quality, I created feature plots using UMI and gene counts. 

It appears that the clustering in EC and pericyte population is due to quality. 

```{r,fig.width=10, fig.height=5}
FeaturePlot(ecpc, features = c("nFeature_RNA", "nCount_RNA"))
```


# Annotate population 

Since sub-clusters in EC population is due to quality, clusters will be annotated at resolution 0.25. 

```{r}
new.cluster.ids <- c("Endothelial_1", "Pericyte", "Endothelial_2", "SMC")
names(new.cluster.ids) <- levels(ecpc_0.25)
ecpc_0.25 <- RenameIdents(ecpc_0.25, new.cluster.ids)
DimPlot(ecpc_0.25, reduction = "umap", label = TRUE, label.size = 8) + NoLegend()
```


```{r}
# Save annotated RDS 
saveRDS(ecpc_0.25, "/Users/serachoi/Documents/Edinburgh/MScProject/Project/data/processed/ecpc04.RDS")
```
