---
title: "06_Luise_SVD"
author: "Sera Choi"
date: "08/06/2021"
output: html_document
---

```{r, message=FALSE, warning=FALSE}
# Load library
library(Seurat)
library(ggplot2)
library(dplyr)
library(SingleCellExperiment)
library(here)
library(ggsci)
library(gtools)
library(Matrix)
library(future)

# Set maximum size of the object 
options(future.global.maxSize = 16000*1024^2)
```

# SVD dataset

In the previous analysis, cluster 2 could not be identified as 50% of cells in cluster 2 belonged to an individual whose death was unknown. However, DEGs indicate that the cells in cluster 2 may be inflammed or damanged. So, here I tried to integrate cells from individuals who were diagnosed with SVD/neurodisease to see if those cells integrate with cluster 2.

Data (GSE157827) contains patients with Alzheimer's disesase and 4 individuals with:

* NC3: microinfact in putamen
* NC16: mild Cerebral amyloid angiopathy
* NC18: mild Cerebral amyloid angiopathy
* NC17: mild SVD

## Load data
```{r}
svd <- readRDS("/Users/serachoi/Documents/Edinburgh/MScProject/Project/data/processed/svd.RDS")
```

## Quality Control

The study removed nuclei with:

* nFeatures_RNA (gene) < 200
* nCount_RNA (library size) > 20,000
* percent.mt > 20%

I followed their QC thresholds. 
NC16 appears to have high mito percentage and low library size compared to other samples. But since I know each sample contains very little endothelial cells, I decided to keep NC16 for now.  

```{r,fig.width= 10, fig.height= 8}
svd[["percent.mt"]] <- PercentageFeatureSet(svd, pattern = "^MT-")
metadata <- svd@meta.data
VlnPlot(svd, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3, pt.size = FALSE)
plot1 <- FeatureScatter(svd, feature1 = "nCount_RNA", feature2 = "percent.mt")
plot2 <- FeatureScatter(svd, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")
plot1 + plot2


# Remove low-quality cells 
svd <- subset(svd, subset = nFeature_RNA > 200 & nFeature_RNA < 20000 & percent.mt < 20)

VlnPlot(svd, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3, pt.size = FALSE)
plot1 <- FeatureScatter(svd, feature1 = "nCount_RNA", feature2 = "percent.mt")
plot2 <- FeatureScatter(svd, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")
plot1 + plot2

```

## Normalisation

The study simple used Seurat basic normalisation method. 

```{r}
svd <- NormalizeData(svd, normalization.method = "LogNormalize", scale.factor = 10000)
svd <- NormalizeData(svd)
```

## Feature Selection and Scaling 
```{r, message=FALSE}
svd <- FindVariableFeatures(svd, selection.method = "vst", nfeatures = 2000)
all.genes <- rownames(svd)
svd <- ScaleData(svd, features = all.genes)
```

## Linear Dimensional reduction 
```{r, message=FALSE}
svd <- RunPCA(svd, features = VariableFeatures(object = svd))
DimPlot(svd, reduction = "pca")
ElbowPlot(svd, ndims = 50)
```

## Visualisation

The study used 20 PCs and cluster resolution = 1. 
Shows clustering due to sample differences. So integration is required. 

```{r, message=FALSE, warning=FALSE}
svd <- RunUMAP(svd, dims = 1:20)
DimPlot(svd, reduction='umap',group.by = "orig.ident")
```

## Integration

After integration, batch effect disappears and UMAP representation resembles that in the study. 

```{r, message=FALSE, warning=FALSE}
# Split Seurat object into 4 objects corresponding to each sample 
svd.list <- SplitObject(svd, split.by = "orig.ident")

# Normalise and identify HVGs for each dataset independently
svd.list <- lapply(X = svd.list, FUN = function(x) {
    x <- NormalizeData(x)
    x <- FindVariableFeatures(x, selection.method = "vst", nfeatures = 2000)
})

# select features that are repeatedly variable across datasets for integration
features <- SelectIntegrationFeatures(object.list = svd.list)
svd.anchors <- FindIntegrationAnchors(object.list = svd.list, anchor.features = features, dims = 1:20)
svd.combined <- IntegrateData(anchorset = svd.anchors, dims=1:20)

DefaultAssay(svd.combined) <- "integrated"

# Run the standard workflow for visualization and clustering
# Feature Selection and Scaling 
svd.combined <- FindVariableFeatures(svd.combined, selection.method = "vst", nfeatures = 2000)
all.genes <- rownames(svd.combined)
svd.combined <- ScaleData(svd.combined, features = all.genes)

# Linear dimensional reduction 
svd.combined <- RunPCA(svd.combined, features = VariableFeatures(object = svd.combined))
ElbowPlot(svd.combined, ndims = 50)

# Visualisation
svd.combined <- RunUMAP(svd.combined, dims = 1:20)
DimPlot(svd.combined, reduction = "umap", group.by = "orig.ident")
```

At resolution 0.1, cluster 12 appears to express EC markers. I used EC markers the study used to identify endothelial cells. 

```{r, fig.width= 15, fig.height= 13}
FeaturePlot(svd.combined, features= c("CLDN5",
                                      "FLT1",
                                      "ABCB1"))

```

## Subset Endo population 

```{r}
Idents(svd.combined) <- "RNA_snn_res.0.1"
svd_ec <- subset(svd.combined, ident=12)
saveRDS(svd_ec, "/Users/serachoi/Documents/Edinburgh/MScProject/Project/data/processed/svd_ec.RDS")
```

## Feature Selection and Scaling 
```{r, message=FALSE}
svd_ec <- readRDS(here("/Users/serachoi/Documents/Edinburgh/MScProject/Project/data/processed/svd_ec.RDS")) 
svd_ec <- FindVariableFeatures(svd_ec, selection.method = "vst", nfeatures = 2000)
all.genes <- rownames(svd_ec)
svd_ec <- ScaleData(svd_ec, features = all.genes)
```

## Linear Dimensional reduction 
```{r, message=FALSE}
svd_ec <- RunPCA(svd_ec, features = VariableFeatures(object = svd_ec))
DimPlot(svd_ec, reduction = "pca")
ElbowPlot(svd_ec, ndims = 50)
```

## Visualisation

Similar to Luise's dataset, two sub-populations appear corresponding to ECs and pericytes. 

```{r, message=FALSE}
# Clustering
svd_ec <- FindNeighbors(svd_ec, dims = 1:20)
svd_ec <- FindClusters(svd_ec, resolution = seq(from = 0.1, to = 1, by = 0.1))
svd_ec <- RunUMAP(svd_ec, dims = 1:20)

plot_list_func(svd_ec, col_pattern = "integrated_snn_res.")
```

### EC markers

```{r, fig.width= 15, fig.height= 13}
FeaturePlot(svd_ec, features= c("CLDN5", 
                               "VWF", 
                               "FLT1", 
                               "ANO2", 
                               "ST6GALNAC3", 
                               "ELOVL7", 
                               "ABCB1"))
```

## Pericyte markers 

```{r, fig.width= 15, fig.height= 13}
FeaturePlot(svd_ec, features= c("PDGFRB", 
                              "NOTCH3", 
                              "PRKG1", 
                              "LAMA2", 
                              "GPC5", 
                              "PDE7B",
                              "DLC1"))
```

## Subset ECs from SVD dataset

At resoltution 0.2, cluster 0 appears to correpond to ECs while cluster 1, pericytes. So, I subset for ECs.

```{r}
Idents(svd_ec) <- "integrated_snn_res.0.2"
svd_ec <- subset(svd_ec, ident=0)
saveRDS(svd_ec, "/Users/serachoi/Documents/Edinburgh/MScProject/Project/data/processed/svd_ec.RDS")
```

# Integration with Luise's data

```{r}
ec <- readRDS(here("/Users/serachoi/Documents/Edinburgh/MScProject/Project/data/processed/ec05.RDS")) 
ec[["orig.ident"]] <- "norm"
svd_ec[["orig.ident"]] <- "svd"
ec.comb <- merge(x = ec, y = svd_ec, merge.data = TRUE)
metadata <- ec.comb@meta.data

ec.list <- SplitObject(ec.comb, split.by = "orig.ident")

# Normalisation and feature selection
for (i in 1:length(ec.list)) {
    ec.list[[i]] <- NormalizeData(ec.list[[i]], verbose = FALSE)
    ec.list[[i]] <- FindVariableFeatures(ec.list[[i]], selection.method = "vst", 
        nfeatures = 2000, verbose = FALSE)
}

# select features that are repeatedly variable across datasets for integration
features <- SelectIntegrationFeatures(object.list = ec.list)
ec.anchors <- FindIntegrationAnchors(object.list = ec.list, anchor.features = features, dims = 1:20)
ec.comb <- IntegrateData(anchorset = ec.anchors, dims=1:20)

DefaultAssay(ec.comb) <- "integrated"

# Run the standard workflow for visualization and clustering
# Feature Selection and Scaling 
ec.comb <- FindVariableFeatures(ec.comb, selection.method = "vst", nfeatures = 2000)
all.genes <- rownames(ec.comb)
ec.comb <- ScaleData(ec.comb, features = all.genes)

# Linear dimensional reduction 
ec.comb <- RunPCA(ec.comb, features = VariableFeatures(object = ec.comb))
ElbowPlot(ec.comb, ndims = 50)

# Visualisation
ec.comb <- RunUMAP(ec.comb, dims = 1:20)
DimPlot(ec.comb, reduction = "umap", group.by = "orig.ident")
```



```



