---
title: "07_age_Luise"
author: "Sera Choi"
date: "13/06/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE, warning=FALSE}
# Load library
library(Seurat)
library(ggplot2)
library(dplyr)
library(here)
library("scales")
library(ggsci)
library(gtools)

# Load data
ec <- readRDS(here("/Users/serachoi/Documents/Edinburgh/MScProject/Project/data/processed/ec05.RDS")) 
```

# Differential gene expression with age

This section performs DE analysis on ECs of young and old brain tissues. 

Precious study on mouse hippocampus ageing (Brain Endothelial Cells Are Exquisite Sensors of Age- Related Circulatory Cues, 2020) has shown that majority of DEGs were increased upon ageing than decreased. 

## Enhanced genes in Old 

Overall, 7 genes are expressed higher in old than young that satisfy p-value < 0.05. Of these, only two genes satisfy the pct.1, pct.2 and adjusted p-value thresholds (ACER2, FGD6)

* SLC7A5: An amino acid transporter in BBB, found to be upregulated in breast, prostate and lung cancer 
* ACER2: Alkaline ceramidase 2 enhanced in capillary ECs of old mouse hippocampus tissue (Brain Endothelial Cells Are Exquisite Sensors of Age- Related Circulatory Cues, 2020). ACER2 promotes tumour angiogenesis (expansion of capillary network to supply tumour with nutrients and oxygen) (https://doi.org/10.1096/fasebj.25.1_supplement.739.10)
* FGD6: Involved in EC apoptosis during vascular remodeling (infammatory response?)
* TACC1: involved in fibrinolysis, extracellular matrix processing 

Here I plotted Violin plot of top 10 enhanced genes in cells from Old individuals. 

```{r,fig.width= 15, fig.height= 13, message=FALSE, warning=FALSE}
Idents(ec) <- "AgeGroup"

# enhanced genes in Old individuals
old.markers <- FindMarkers(ec,
                           ident.1 = "Old",
                           ident.2 = "Young",
                           only.pos = TRUE, 
                           min.pct = 0.25, 
                           logfc.threshold = 0.25,
                           test.use = "MAST")

old.markers <- subset(old.markers, old.markers$p_val_adj < 0.05)

old.markers_sig <- subset(old.markers,
                          old.markers$pct.1 > 0.25 &
                          old.markers$pct.2 < 0.25 )
old.markers
old.markers_sig

VlnPlot(ec, features = rownames(old.markers), group.by = "AgeGroup")
```

# Young-enhanced genes 

None are enahnced. Three genes are identified in Young that satisfy p-value < 0.05. 

* ABCG2 was enahnced in aged cells in mouse hippocampus. 

```{r,fig.width= 15, fig.height= 7, message=FALSE, warning=FALSE}
Idents(ec) <- "AgeGroup"

# enhanced genes in Young individuals
young.markers <- FindMarkers(ec,
                           ident.1 = "Young",
                           ident.2 = "Old",
                           only.pos = TRUE, 
                           min.pct = 0.25, 
                           logfc.threshold = 0.25,
                           test.use = "MAST")

young.markers <- subset(young.markers,
                            young.markers$p_val_adj < 0.05)
young.markers
VlnPlot(ec, features = rownames(young.markers), group.by = "AgeGroup")
```

# Dot Plot

```{r, fig.width= 13}
Idents(ec) <- "AgeGroup"
DotPlot(ec, 
        features=c(rownames(old.markers),rownames(young.markers)))
```


# Higher expression in aged cells?

The study on ageing of mouse hippocampus showed that aged cells expressed more genes. Here, I tried checking this. 

```{r}
VlnPlot(ec, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), group.by="AgeGroup",ncol=3 )
```

# Conlcusion

* In this dataset, genes are enhanced in aged cells and none are up-regulated in younger cells (sort of consistent with studies of ageing in mouse hippocampus)
* Up-regulated genes in aged cells are previously known to be associated with cancer, inflammatory response and regeneration of ECMs. (consistent with hallmarks of ageing: neuroinflammation)
* More genes and mito genes expressed in older cells (consistent with other studies)
* Aged capillary has enhanced expression of genes associated with immunity, antigen-processing, matrix assembly and oxidative stress pathways. 



